Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/compiler.tex
Edits to MOTION section.

MOTION supports FOR loops and SIMD operations, so translation from MPC source to MOTION C++ code is relatively straightforward.
% Variable naming
\paragraph{Variable declarations:}
Our compiler performs taint analysis on MPC Source based on the type system in~\secref{sec:syntax}. The analysis is described in the following section;
for the most part, it is a standard taint analysis that propagates that \texttt{shared} qualifiers throughout the routine.
Taint analysis assigns \texttt{shared} or \texttt{plain} qualifiers to all local variables, which is important for code generation.
Our generated C++ uses the following variable-naming scheme: shared variables are named the same as in the MPC Source with the {\sf !} replaced with an underscore (e.g. {\sf sum!2} would be translated to {\sf sum\_2}). Plaintext variables follow the same naming convention as shared variables but are prefixed with {\sf \_MPC\_PLAINTEXT\_}.  The shared representation of constants are named {\sf \_MPC\_CONSTANT\_} followed by the literal constant (e.g. the shared constant 0 would be named {\sf \_MPC\_CONSTANT\_0}).

@@

% Phi nodes and FOR loops
Pseudo PHI nodes are broken into two components: the ``FALSE'' branch which assigns the initial value of the PHI node and the ``TRUE'' branch which assigns the PHI node's back-edge.  The assignment of the false branch occurs right before the PHI node's enclosing loop.  Inside of the PHI node's enclosing loop, a C++ \texttt{if} statement is inserted to only assign the true branch of the PHI node after the first iteration. MPC FOR loops are converted to C++ FOR loops which iterate the loop counter over the specified range.  The loop counter is initialized before any PHI nodes in the loop as they could rely on it; there is no other special handling of FOR loops. \ana{Add the reference to the corresponding figure somewhere in this paragraph.}

@@

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional \texttt{std::vector}s in row-major order. \ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} \ben{MOTION does only support 1-dim SIMD arrays, but since we use helper functions to create and access SIMD arrays this restriction doesn't apply to our representation. There is no reason why these arrays need to be 1-dimensional other than that it slightly simplifies the code generation and avoids some nasty templates on the utility functions.} Whenever a SIMD value is used in an expression, the utility function \texttt{vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. \ana{Add the reference to the ADD example figure here.} Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression. \ana{Is it correct to say that the operators are overloaded? If yes, use that term?} \ben{I don't think that's the right term since SIMD and scalar MOTION values use the same C++ class, so the operator is the same.} Therefore, the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

Similarly, the \texttt{vectorized\_assign()} function assigns a (potentially SIMD) value to a slice of a vectorized array.  This operation cannot be done with a simple subscript as SIMD assignments will update a range of values in the non-vectorized representation.
%\ben{Is the next sentence needed?  They can always check the implementation of vectorized\_assign() if they want to see specifics.}
%The function takes the array being assigned to as a reference to update it in place so that any existing values are not lost.  SIMD values are never directly assigned to non-SIMD values. \ana{Yes, this is implementation-specific, you can remove.}

@@

% vectorized_update()
\ben{Technically this is semantically the same as non-vectorized update but we use a utility function to combine the steps into a single C++ statement - should it still be included?}
\ana{I would leave this in the TR but it will go from the paper. Lots of stuff in Secs. 4 and 5 will go from the paper.}
Updating SIMD arrays is also implemented differently from updating non-vectorized arrays.  Instead of separating the array update from the assignment of the new array, these steps are combined with the \texttt{vectorized\_update()} utility function.  This function operates identically to \texttt{vectorized\_assign()}, however it additionally returns the array after the assignment occurs.  This value is then used for the assignment to the new variable.

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{Yes, that's true. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the \texttt{lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  This function evaluates the expression for each permutation of indices along the dimensions and returns the resulting array in row-major order.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the \texttt{lift()} function which occurs when we are raising an array.  In this case, instead of concatenating the array for each index, we extend the array along all dimensions being raised which are not present in the array already.  For example, when raising an array with dimensions $N \times M$ to an array with dimensions $N \times M \times D$, the input array will simply be extended along the $D$ dimension.  If input array is already correctly sized it will be returned as-is.
%This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).
\ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.} \ben{Rewrote the above paragraph to be less confusing/implementation-specific, please review.}

@@

% drop_dim() implementation
\paragraph{Dropping dimensions:}
Dropping dimensions use the \texttt{drop\_dim()} and \texttt{drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.

@@

% Tons of copies != runtime costs
Due to the SSA translation phase as well as the various conversions to and from SIMD values which our utility functions perform, our generated MOTION code often includes many copies of arrays and scalar values.  These copies do not incur a runtime cost as the arrays simply hold pointers to the underlying shares, so no new shares or gates are created as a result of this copying. Cost in MPC programs is dominated by shares and computation on shares. \ben{Is there anything else to say here?  Should this be placed somewhere else?} \ana{It is clear to me! Will stay in the TR but likely will be cut from the paper.}

>>>>

MOTION supports FOR loops and SIMD operations, so translation from MPC source to MOTION C++ code is relatively straightforward.
% Variable naming
\paragraph{Variable declarations:}
Our compiler performs taint analysis on MPC Source based on the type system in~\secref{sec:syntax}. The analysis is described in the following section;
for the most part, it is a standard taint analysis that propagates the declared \texttt{shared} qualifiers throughout the routine.
Taint analysis assigns \texttt{shared} or \texttt{plain} qualifiers to all local variables, which is important for code generation.
Our generated C++ uses the following variable-naming scheme: shared variables are named the same as in the MPC Source with the {\sf !} replaced with an underscore (e.g. {\sf sum!2} would be translated to {\sf sum\_2}). Plaintext variables follow the same naming convention as shared variables but are prefixed with {\sf \_MPC\_PLAINTEXT\_}.  The shared representation of constants are named {\sf \_MPC\_CONSTANT\_} followed by the literal constant (e.g. the shared constant 0 would be named {\sf \_MPC\_CONSTANT\_0}).

@@

% Phi nodes and FOR loops
Pseudo PHI nodes are broken into two components: the ``FALSE'' branch which assigns the initial value of the PHI node and the ``TRUE'' branch which assigns the PHI node's back-edge.  The assignment of the false branch occurs right before the PHI node's enclosing loop.  Inside of the PHI node's enclosing loop, a C++ \texttt{if} statement is inserted to only assign the true branch of the PHI node after the first iteration. MPC FOR loops are converted to C++ FOR loops which iterate the loop counter over the specified range.  Listing~\ref{tab:motion_translation_for_loop} illustrates. The loop counter is initialized before any PHI nodes in the loop as they could rely on it; there is no other special handling of FOR loops. %\ana{Add the reference to the corresponding figure somewhere in this paragraph.}

@@

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional {\sf std::vector}s in row-major order. %\ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} \ben{MOTION does only support 1-dim SIMD arrays, but since we use helper functions to create and access SIMD arrays this restriction doesn't apply to our representation. There is no reason why these arrays need to be 1-dimensional other than that it slightly simplifies the code generation and avoids some nasty templates on the utility functions.} \ana{Ok! Leave as is.} 
Whenever a SIMD value is used in an expression, the utility function {\sf vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression. 
%\ana{Is it correct to say that the operators are overloaded? If yes, use that term?} \ben{I don't think that's the right term since SIMD and scalar MOTION values use the same C++ class, so the operator is the same.} \ana{Ok, leave as is.}
Therefore, the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

Similarly, the {\sf vectorized\_assign()} function assigns a (potentially SIMD) value to a slice of a vectorized array.  This operation cannot be done with a simple subscript as SIMD assignments will update a range of values in the non-vectorized representation.
%\ben{Is the next sentence needed?  They can always check the implementation of vectorized\_assign() if they want to see specifics.}
%The function takes the array being assigned to as a reference to update it in place so that any existing values are not lost.  SIMD values are never directly assigned to non-SIMD values. \ana{Yes, this is implementation-specific, you can remove.}

@@

% vectorized_update()
%\ben{Technically this is semantically the same as non-vectorized update but we use a utility function to combine the steps into a single C++ statement - should it still be included?}
%\ana{I would leave this in the TR but it will go from the paper. Lots of stuff in Secs. 4 and 5 will go from the paper.} \ana{Leaving in TR, will shorten in paper.}
Updating SIMD arrays is also implemented differently from updating non-vectorized arrays.  Instead of separating the array update from the assignment of the new array, these steps are combined with the {\sf vectorized\_update()} utility function.  This function operates identically to {\sf vectorized\_assign()}, however it additionally returns the array after the assignment occurs.  This value is then used for the assignment to the new variable. Listing~\ref{tab:motion_translation_simd_assignment} illustrates {\sf vectorized\_assign()} and {\sf vectorized\_update()} on the Biometric example.

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{True. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now but leave the comments too because it needs work.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the {\sf lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  This function evaluates the expression for each permutation of indices along the dimensions and returns the resulting array in row-major order.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the {\sf lift()} function which occurs when we are raising an array.  In this case, instead of concatenating the array for each index, we extend the array along all dimensions being raised which are not present in the array already.  For example, when raising an array with dimensions $N \times M$ to an array with dimensions $N \times M \times D$, the input array will simply be extended along the $D$ dimension: {\sf A'[n,m,d] = A[n,m]} for every $d$.  If the input array is already correctly sized it will be returned as-is.
%This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).
%\ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.} \ben{Rewrote the above paragraph to be less confusing/implementation-specific, please review.} \ana{Better, removing comment.}

@@

% drop_dim() implementation
\paragraph{Dropping dimensions:}
Dropping dimensions use the {\sf drop\_dim()} and {\sf drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.

@@

% Tons of copies != runtime costs
Due to the SSA translation phase as well as the conversions to and from SIMD values which our utility functions perform, our generated vectorized MOTION code often includes multiple copies of arrays and scalar values.  These copies do not incur a runtime cost as the arrays simply hold \emph{pointers} to the underlying shares, so no new shares or gates are created as a result of this copying. Cost in MPC programs is dominated by shares and computation on shares. %\ben{Is there anything else to say here?  Should this be placed somewhere else?} \ana{It is clear to me! Will stay in the TR but likely will be cut from the paper.}
